---
title: "ENV S 193DS Final"
author: "Sumaiya Alvi"
date: "March 17, 2026"
format: 
  html:
    toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show up
---

[link to Github repository](https://github.com/sumaiyaalvi/ENVS-193DS_final.git)

# Set up

```{r}
#| label: load packages and data
#| message: false

library(tidyverse) #general use
library(here) #file organization
library(janitor) #cleaning data frames
library(readxl) #reading excel files

kelp <- read_csv(
  here( 
    "data", #data folder
    "Annual_Kelp_All_Years_20250903.csv" #kelp data 
  ))

screentime <- read_csv(
  here( 
    "data", #data folder
    "screentime.csv" #personal data 
  ))

mopl <- read_csv(
  here( 
    "data", #data folder
    "MOPL_nest-site_selection_Duchardt_et_al._2019.csv" #mountain plover data 
  )) 
```

# Problem 1. Research Writing

## a. Transparent statistical methods

In **part 1**, they used a univariate general linear model to examine the relationship between precipitation and flooded wetland area.

In **part 2**, they used a one-way analysis of variance (ANOVA) to test whether flooded wetland area differed across the water year classifications.

## b. Figure needed

A useful figure for the one-way ANOVA would be a boxplot showing flooded wetland area across water-year classifications. The x-axis should show the water year classification categories (wet, above normal, below normal, dry, and critical drought). The y-axis should show flooded wetland area (m²). Each boxplot would represent the distribution of flooded wetland area within each classification. Individual data points could also be added as jittered points to show the spread of the observations for each category. This figure would help readers visually compare flooded wetland area across the different water year classifications and better understand the results of the ANOVA test.

## c. Suggestions for rewriting

We did not find strong evidence that precipitation predicts flooded wetland area in the San Joaquin River Delta (univariate general linear model; test statistic = unknown; $\alpha$ = significance level; $p$ = 0.11). Flooded wetland area also did not appear to differ across water year classifications (wet, above normal, below normal, dry, critical drought), which we visualized with a boxplot comparing flooded wetland area across categories (one-way ANOVA; test statistic = unknown; df = unknown; $\alpha$ = significance level; $p$ = 0.12).

## d. Interpretation

One additional variable that could influence flooded wetland area is **water salinity**, which is a continuous variable. Higher salinity levels may make water less suitable for maintaining wetland habitat, so managers may choose not to flood or maintain wetlands when salinity is high, which could reduce the flooded area.

# Problem 2. Data visualization

## a. Cleaning and summarizing

```{r}
kelp_clean <- kelp |> 
  # function 3:
  # make all column names lowercase and consistent
  clean_names() |> 
  # function 6
  # keep only the relevant observations
  filter(site %in% c("IVEE", "NAPL", "CARP")) |> 
  # function 1
  # remove the specific observations that meet all 3 of these criteria
  filter(!(date == "2000-09-22" & transect == "6" & quad == "40")) |> 
  # function 8
  # create new site_name column with full site names 
  mutate(site_name = case_when(
  site == "NAPL" ~ "Naples",
  site == "IVEE" ~ "Isla Vista",
  site == "CARP" ~ "Carpinteria"
  )) |> 
  # function 2
  # convert site_name to a factor and set the 3 factor levels
  mutate(site_name = as_factor(site_name),
         site_name = fct_relevel(site_name,
                               "Naples", "Isla Vista", "Carpinteria")) |> 
  # function 5
  # group data by site_name and year
  group_by(site_name, year) |> 
  # function 7
  # calculate the mean number of fronds for each site-year combination
  summarize(mean_fronds = mean(fronds)) |> 
  # function 4
  # remove the grouping by site_name and year
  ungroup()

kelp_clean %>%
  slice_sample(n = 5)

str(kelp_clean)
```










